% paper/phase_i.tex — Workshop paper draft
% Title: Porting ZGC to CHERI/Morello — design choices for moving GC
%        on immutable capabilities
%
% Target venues: VMIL @ SPLASH, ICOOLPS @ ECOOP, MARC, DSbD annual.
% First arXiv preprint goes up on submission.

\documentclass[acmsmall,nonacm]{acmart}

\usepackage{listings}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{hyperref}

\title{Porting ZGC to CHERI/Morello: \\
       Design Choices for Moving GC on Immutable Capabilities}
\author{[Anonymised for review]}
\affiliation{\institution{[Anonymised]}\country{}}
\email{}

\begin{abstract}
ZGC is OpenJDK's flagship low-latency concurrent moving garbage collector.
Its design rests on two structural features—colored pointers in
unused virtual address bits, and a four-way virtual mapping of the heap—
that collide with the immutability and bounds-discipline of CHERI
capabilities. As a result, MOJO's existing port of OpenJDK 17 to
Morello/CheriBSD covers only the non-moving (Epsilon, Serial) and
region-based (G1) collectors. We present a port of ZGC to CHERI/Morello,
discuss the three design choices forced on us by CHERI's pointer
discipline, and report initial functional results on the Morello FVP.
The port preserves ZGC's pause-time properties while introducing a
side-table color encoding and a per-region capability derivation
strategy. This is the substrate for our follow-up work on a
CHERI-native hardware load barrier, where the cap-load tag itself
subsumes ZGC's software color check.
\end{abstract}

\begin{document}
\maketitle

% ---------------------------------------------------------------------
\section{Introduction}
\label{sec:intro}

Modern Java workloads demand garbage collectors that bound pause time to
the sub-millisecond range. ZGC~\cite{zgc-toplas} achieves this by
concurrently relocating objects in the background while mutators run, at
the cost of a software \emph{load barrier} executed on every reference
load—typically an \texttt{AND} against a precomputed ``bad mask''
followed by a conditional branch into a slow path. On pointer-chasing
workloads this overhead can consume a non-trivial fraction of CPU.

CHERI~\cite{cheri-isa-v9} provides hardware-checked, tagged pointers in
which every load is already validated by the LSU at no software cost.
The natural cross-domain idea—use CHERI's hardware tag as ZGC's load
barrier—has not been built. Existing CHERI work on memory management
(CHERIvoke~\cite{cherivoke}, Cornucopia~\cite{cornucopia}, Cornucopia
Reloaded~\cite{cornucopia-reloaded}) addresses \emph{revocation} of
capabilities for C/C++ temporal safety; it does not address the
\emph{forwarding} needed by a Java-class moving GC, where every
reference to a relocated object must resolve to the new location.

Before that idea can be tested, ZGC has to exist on CHERI. The MOJO
project~\cite{mojo-jvm} has ported OpenJDK 17's non-moving and
region-based collectors to CheriBSD on Morello, but not ZGC—because
ZGC's two defining optimisations, colored pointers and multi-mapping,
both conflict with CHERI's pointer format.

This paper is the engineering substrate. We describe the design
choices needed to port ZGC to CHERI semantically (without yet
exploiting CHERI as a hardware barrier), implement them in a fork of
OpenJDK 17u, and report initial functional results.

% ---------------------------------------------------------------------
\section{Background}
\label{sec:bg}

\subsection{ZGC's pointer encoding}

ZGC stores GC state in unused virtual address bits of every
\texttt{oop}~\cite{zgc-toplas}. Four colors—\texttt{marked0},
\texttt{marked1}, \texttt{remapped}, and \texttt{finalizable}—occupy
the top bits. The fast-path load barrier checks the loaded oop's color
against a current bad mask in a global; if any bad bit is set, the
slow path consults the forwarding table and self-heals the load site.

Color flipping at GC phase boundaries is fast because the four colors
map to four virtual mappings of the same physical heap. The hardware
MMU does the disambiguation; only the bad mask changes between phases.

\subsection{CHERI's pointer discipline}

A Morello capability is 128 bits: a 64-bit address plus bounds,
permissions, an \texttt{otype}, and an out-of-band tag controlled by
hardware. Three properties matter for ZGC:

\begin{itemize}
\item \textbf{Tag is immutable per cap.} Capabilities cannot be rewritten
      in place; you can derive narrower caps from wider ones, but you
      cannot move a cap's target.
\item \textbf{High address bits are real.} The address must lie within
      the cap's bounds; smuggling GC state into the top bits would
      either escape the bounds or rewrite into the wrong object.
\item \textbf{Tagged loads need cap-aware instructions.} Loading a
      memory location with a non-cap instruction discards the tag,
      breaking subsequent dereferences.
\end{itemize}

\subsection{Prior art on CHERI moving runtimes}

Cornucopia Reloaded~\cite{cornucopia-reloaded} achieves near-zero pause
times for capability revocation on Morello by using per-page hardware
load barriers. The mechanism is well-suited to \emph{invalidating}
stale caps, but it does not address forwarding: revoked caps go to a
trap, not to a new target. MOJO~\cite{mojo-jvm} ports the non-moving
and region-based OpenJDK collectors, deferring ZGC and Shenandoah.

% ---------------------------------------------------------------------
\section{Design choices}
\label{sec:design}

We identify three sub-problems forced by CHERI on a ZGC port and
discuss the design space for each.

\subsection{Colored-pointer encoding}
\label{sec:design:color}

ZGC's high-bit color encoding cannot survive CHERI's address
discipline. Three alternatives:

\paragraph{(a) Side-table.} Per-object color in a shadow bitmap addressed
by object address. Costs one extra load per barrier; that load is
L1-hot in steady state. Color flip becomes a region scan. No change to
cap format.

\paragraph{(b) Cap permissions.} Re-purpose two cap permission bits as
color. Hardware-checked per load (free). Cost: every color flip mints
new caps for every reference—the immutability problem in miniature.

\paragraph{(c) Sealed caps / \texttt{otype}.} Sealed capabilities cannot
be dereferenced, defeating the purpose of in-flight pointers.

We choose (a). It also becomes the substrate for our follow-up Phase
2 work, where Cornucopia Reloaded's per-page revocation bitmap takes
the side-table's role.

\subsection{Multi-mapping}
\label{sec:design:multimap}

Once color lives in a side-table, multi-mapping becomes redundant. We
collapse ZGC's four virtual mappings to a single mapping plus the
side-table from \S\ref{sec:design:color}. We lose the cheap color flip
(a global bad-mask update is replaced by a side-table sweep), but
color flips happen at most once per GC cycle.

\subsection{Forwarding table}
\label{sec:design:fwd}

ZGC's forwarding table maps \texttt{from\_addr -> to\_addr}. On CHERI
the destination must be a real capability with bounds. We adopt
per-region cap roots: each ZGC region installs a root cap covering
its extent, and forwarding-table entries are derived from the
destination region's root via \texttt{csetbounds}. Lookup pays one
extra indirection (region table) in the slow path.

% ---------------------------------------------------------------------
\section{Implementation}
\label{sec:impl}

The port is open-source at \href{https://github.com/TBD/stopless-java-gc}
{https://github.com/TBD/stopless-java-gc}. The repository is structured
to keep our contribution boundary visible:

\begin{itemize}
\item \texttt{src/cap\_runtime/}: \mbox{$\sim$2.5\,kLOC} of net-new C++
      providing the side-table, forwarding table, and (Phase 2)
      revocation glue. Linked into HotSpot via a single hook patch.
\item \texttt{patches/openjdk-jdk17/}: \mbox{$\sim$2\,kLOC} of in-place
      patches against OpenJDK 17u's ZGC source, applied
      idempotently by \texttt{scripts/apply\_patches.sh}.
\item \texttt{src/cpu/morello/gc/z/}: \mbox{$\sim$2\,kLOC} new HotSpot
      CPU subdirectory mirroring \texttt{cpu/aarch64/gc/z/} with
      Morello capability-mode load barrier asm.
\end{itemize}

Third-party source is never committed; \texttt{scripts/bootstrap.sh}
clones OpenJDK 17u, MOJO patches, CheriBSD, and the CHERI toolchain
into a gitignored \texttt{third\_party/} directory pinned by SHA.

% ---------------------------------------------------------------------
\section{Evaluation}
\label{sec:eval}

\subsection{Setup}

[Placeholder: Morello FVP cycle-approximate; baseline = MOJO G1 on
the same FVP; comparison = unported OpenJDK 17u on x86\_64 (anchors
absolute numbers).]

\subsection{Functional smoke}

[Placeholder: DaCapo \texttt{h2}, \texttt{pmd} run to completion on
Morello FVP with \texttt{-XX:+UseZGC}, no unexpected capability traps.]

\subsection{Microbenchmarks}

[Placeholder: bare cap-load vs.\ cap-load + side-table consultation,
isolated by \texttt{tests/bench/microbench\_pointer\_chase}.]

% ---------------------------------------------------------------------
\section{Discussion and limitations}
\label{sec:disc}

The side-table approach trades a one-cycle hardware check for two
loads (cap-load plus side-table consult). Phase 2 of this project
recovers the hardware check by using Cornucopia Reloaded's per-page
revocation bit as the side-table; this paper does not address Phase 2.

We have not exercised C2 cap-awareness; current results use C1 + the
interpreter. Production-grade C2 emission of cap instructions remains
future work.

% ---------------------------------------------------------------------
\section{Related work}
\label{sec:related}

CHERIvoke~\cite{cherivoke}, Cornucopia~\cite{cornucopia}, and
Cornucopia Reloaded~\cite{cornucopia-reloaded} target C/C++ temporal
safety via capability revocation. MOJO~\cite{mojo-jvm} ports
non-moving OpenJDK collectors to CheriBSD/Morello. C4~\cite{c4} and
Shenandoah~\cite{shenandoah} pioneered concurrent compaction with
software read barriers. ZGC~\cite{zgc-toplas} is our base GC.

% ---------------------------------------------------------------------
\section{Conclusion}
\label{sec:conc}

ZGC's optimisations are the product of an era when unused virtual
address bits were free. On CHERI they are not. We have ported ZGC by
moving color into a side-table and per-region capability derivation
into the forwarding table, preserving ZGC's pause-time properties at
the cost of one extra load per barrier. This port is the foundation
for a CHERI-native hardware barrier that we describe in a follow-up.

% ---------------------------------------------------------------------
\paragraph{Acknowledgements.}
This work was implemented with substantial assistance from Claude Opus
4.7 via Claude Code (Anthropic). All design decisions, evaluations,
and the final manuscript were authored by humans; the LLM acted as a
literature-review accelerator, design-space enumerator, and code
generator under human direction. See \texttt{docs/04\_risk\_register.md}
in the project repository for a detailed account of what the model
did and did not contribute.

\bibliographystyle{ACM-Reference-Format}
\bibliography{refs}

\end{document}
